### 2014-04-22 docs/paper/paper.tex
system block diagram
dv double buffer diagram

\design{} is a custom digital hardware design that implements the population-mode NEF in real-time (1~ms timestep).
% FIXME figure: block diagram
The encoder units perform the function of encoding (), % FIXME NEF equation
while the population units perform the decoding () % FIXME NEF equation
and write the decoded values to the decoded value buffers
to be used as inputs to other populations or as outputs from the network.
Each population of neurons is assigned to a population unit, which is able to simulate up to 1024 populations
having either one or two encoded inputs each. These inputs are calculated by the encoder units,
of which there are two for a one-dimensional population unit and four for a two-dimensional population unit;
each encoder unit can be associated with a different first-order filter to simulate post-synaptic filtering.
For example, one of the two encoders on a one-dimensional population unit can be programmed with
a small time constant to model fast synapses (e.g. AMPA and GABA$_A$), while the second can be programmed
with a large time constant to model slow synapses (e.g. NMDA and GABA$_B$).
Each population unit performs time-multiplexing and swaps in different sets of decoders
in order to run more populations on the same hardware. The principal components are fixed and must be shared
between all populations on an individual population unit.
Each population unit writes to a fixed set of decoded value buffers, but a shared interconnect
allows the encoder units to perform random accesses to these buffers, which allows
inputs to a population to come from any combination of decoded values. % FIXME more detail
A number of decoded value buffers are reserved for external inputs to the network,
which can be sent from the controlling computer system over an Ethernet connection.
The output channels can also be programmed to send decoded values back to the computer over Ethernet.

@@

Encoded values are passed from the encoder units to the population units. 
Population units are either one-dimensional or two-dimensional.
This means the groups of neurons they model have firing rates that are static functions of either a one- or two-dimensional represented variable.
Both types of population unit receive input from two encoder units in each input dimension;
for each dimension, the encoded values are passed through a first-order filter which performs post-synaptic filtering, % FIXME check this
and then summed. The value resulting from this operation is
truncated to the most significant 10 bits and is then used as input to a set of lookup tables,
implemented as a bank of synchronous RAM elements. SRAMs are used
because the contents of these tables are not known until simulation-time and therefore the lookups cannot be implemented as hardware logic functions.
Each lookup table contains the 12-bit fixed-point values of points sampled from 
%the principal components shared by all populations of neurons
%simulated on that population unit. 
principal components of the spike rates of neurons in a population, as a function of the variable that the population encodes.
As described further below
%(Section FIXME)
, all populations on a population unit must share principal components.
For one-dimensional population units, there are seven such lookup tables, and each one contains
1024 values sampled at equally-spaced points on the open interval $(-2, 2)$. For two-dimensional population units,
there are fifteen lookup tables, each containing 1024 values sampled at equally-spaced points on a square of width 4
centered at $(0, 0)$. The resulting sample space is a 32x32 grid.
These numbers of principal components account for most of the variation in spike rates across 1D and 2D populations.
% FIXME find out exactly how much of this variation they account for over random populations with a variety of parameters, and mention this.
To increase the accuracy of this representation, each two-dimensional population unit performs bilinear interpolation
using the least significant five bits of each input variable to determine the distance between sample points. 
The extra resources used to perform this operation are modest
and significantly reduce the amount of memory used for two-dimensional populations while maintaining a good degree of accuracy of representation.

@@

Decoded values calculated from population units are stored in decoded value buffers, which are fairly large memory banks that can each
store 2048 decoded values. 
Because each population unit produces 4096 decoded values (1024 DVs from each of four decoder vectors),
this means that a population unit must be attached to the write ports of two separate decoded value buffers.
The buffers themselves expose a read-only interface, accessed by the encoders to read decoded values from the
previous timestep, and a write-only interface, accessed by the population units when new decoded values are calculated.
Internally, each interface is connected to a separate dual-port RAM; the connections are swapped between RAMs
at the end of each timestep in order to implement double-buffering, so that it is not possible for encoders to read a
partially-updated set of decoded values, where some are from the previous timestep and some are freshly written during the current timestep.
The reason for choosing a size of 2048 is because of this implementation detail; on the particular family of hardware
that was used to implement the prototype, the largest single-component 12-bit dual-port RAM available has space for 1024 words,
and two separate RAM components are necessary in order to implement the double buffer, hence a total size of 2048 words of memory.
% FIXME DV double buffer diagram in here somewhere

>>>>

\design{} is a custom digital hardware design that implements the population-mode NEF in real-time (1~ms timestep).
A block diagram of the design is shown in Figure~\ref{fig:system}.
The encoder units perform the function of encoding (), % FIXME NEF equation
while the population units perform the decoding () % FIXME NEF equation
and write the decoded values to the decoded value buffers
to be used as inputs to other populations or as outputs from the network.
Each population of neurons is assigned to a population unit, which is able to simulate up to 1024 populations
having either one or two encoded inputs each. These inputs are calculated by the encoder units,
of which there are two for a one-dimensional population unit and four for a two-dimensional population unit;
each encoder unit can be associated with a different first-order filter to simulate post-synaptic filtering.
For example, one of the two encoders on a one-dimensional population unit can be programmed with
a small time constant to model fast synapses (e.g. AMPA and GABA$_A$), while the second can be programmed
with a large time constant to model slow synapses (e.g. NMDA and GABA$_B$).
Each population unit performs time-multiplexing and swaps in different sets of decoders
in order to run more populations on the same hardware. The principal components are fixed and must be shared
between all populations on an individual population unit.
Each population unit writes to a fixed set of decoded value buffers, but a shared interconnect
allows the encoder units to perform random accesses to these buffers.
This allows effective all-to-all communication between populations while reducing the number of 
connections that must be made between population units.
A number of decoded value buffers are reserved for external inputs to the network,
which can be sent from the controlling computer system over an Ethernet connection.
The output channels can also be programmed to send decoded values back to the computer over Ethernet.

\includegraphics[width=6in]{system-block-diagram.eps}

\caption{Block diagram of the \design{} hardware design.}

\label{fig:system}
\end{figure}

@@

Encoded values are passed from the encoder units to the population units. 
Population units are either one-dimensional or two-dimensional.
This means the groups of neurons they model have firing rates that are static functions of either a one- or two-dimensional represented variable.
Both types of population unit receive input from two encoder units in each input dimension;
for each dimension, the encoded values are passed through a first-order filter which performs post-synaptic filtering,
and are then summed. The value resulting from this operation is
truncated to the most significant 10 bits and is then used as input to a set of lookup tables,
implemented as a bank of synchronous RAM elements. SRAMs are used
because the contents of these tables are not known until simulation-time and therefore the lookups cannot be implemented as hardware logic functions.
Each lookup table contains the 12-bit fixed-point values of points sampled from 
principal components of the spike rates of neurons in a population, as a function of the variable that the population encodes.
In order to reduce the amount of memory used by each population, all populations on a population unit must share principal components.
For one-dimensional population units, there are seven such lookup tables, and each one contains
1024 values sampled at equally-spaced points on the open interval $(-2, 2)$. For two-dimensional population units,
there are fifteen lookup tables, each containing 1024 values sampled at equally-spaced points on a square of width 4
centered at $(0, 0)$. The resulting sample space is a 32x32 grid.
These numbers of principal components account for most of the variation in spike rates across 1D and 2D populations.
% FIXME find out exactly how much of this variation they account for over random populations with a variety of parameters, and mention this.
To increase the accuracy of this representation, each two-dimensional population unit performs bilinear interpolation
using the least significant five bits of each input variable to determine the distance between sample points. 
The extra resources used to perform this operation are modest
and significantly reduce the amount of memory used for two-dimensional populations while maintaining a good degree of accuracy of representation.

@@

Decoded values calculated from population units are stored in decoded value buffers, which are fairly large memory banks that can each
store 2048 decoded values. 
Because each population unit produces 4096 decoded values (1024 DVs from each of four decoder vectors),
this means that a population unit must be attached to the write ports of two separate decoded value buffers.
The buffers themselves expose a read-only interface, accessed by the encoders to read decoded values from the
previous timestep, and a write-only interface, accessed by the population units when new decoded values are calculated.
Internally, each interface is connected to a separate dual-port RAM; the connections are swapped between RAMs
at the end of each timestep in order to implement double-buffering, so that it is not possible for encoders to read a
partially-updated set of decoded values, where some are from the previous timestep and some are freshly written during the current timestep.
A diagram of this component is shown in Figure~\ref{fig:dvdoublebuffer}.
The reason for choosing a size of 2048 is because of this implementation detail; on the particular family of hardware
that was used to implement the prototype, the largest single-component 12-bit dual-port RAM available has space for 1024 words,
and two separate RAM components are necessary in order to implement the double buffer, hence a total size of 2048 words of memory.

\includegraphics[width=6.0in]{dv-double-buffer-diagram.eps}

\caption{Block diagram of the decoded value double-buffer component.}
\label{fig:dvdoublebuffer}
\end{figure}
